Making Sequential Programs Deterministic

### 2019-03-28 design.tex
todo: redraw figure
todo: type welformedness consistency
todo: resolve -> instantiate

\begin{wrapfigure}{R}{0.3\textwidth}
    \begin{center}
        \includegraphics[scale=0.5]{determinism}
    \end{center}
    \caption{Determinism type qualifier hierarchy}
    \label{fig:determinism-hierarchy}
\todo{Drop the at-signs (@).}
\todo{The text is weirdly small for the box.
  Probably redraw this.}
\end{wrapfigure}

@@

\Cref{type-validity} shows the type validity\todo{The caption says
  ``type well formedness''.  Choose one term and use it consistently
  throughout.}
rules of \ourTypeSystem. 
As stated in the rule \rulename{noncollection}, the types \<Det> and \<NonDet>
can be written on any non-collection Java basetype.

@@

When a user writes a polymorphic qualifier on a method signature or a type parameter,
it indicates that it could resolve\todo{Should all occurrences of ``resolve'' be ``can be instantiated''} to any type in the type system depending on how it is used.
In \theDeterminismChecker, we define a polymorphic qualifier \<PolyDet>.
One of the most common locations for a polymorphic qualifier is at a method signature.
Consider the following declaration of addition:
\begin{verbatim}
@PolyDet int plus(@PolyDet int a, @PolyDet int b) { return a+b; }
\end{verbatim}
\todo{It has the standard meaning.  There are two ways to interpret this.}
There are two valid instantiations for this declaration:
\begin{Verbatim}
  @NonDet int plus(@NonDet int a, @NonDet int b) { return a+b; }
  @Det int plus(@Det int a, @Det int b) { return a+b; }
\end{Verbatim}
This indicates that method \<foo> can be called with arguments having any of the \codeid{Det or NonDet} type qualifiers.
%   (\<@OrderNonDet> is not allowed because it is invalid on primitive types).
 \<PolyDet> resolves to the least upper bound of
the actual types on arguments. For instance, if this method is called as \codeid{foo(@Det String arg1, @NonDet boolean arg2)}, \theDeterminismChecker resolves the method declaration as \codeid{@NonDet int foo(@NonDet String param1, @NonDet boolean param2)}
causing the return type to have the type qualifier \<NonDet>.

@@

Recall that \theDeterminismChecker does not permit \<OrderNonDet> on a type
that isn't a collection.
As a consequence, resolution of polymorphic type qualifiers at method signatures could have undesirable effects.
Suppose a method declaration is annotated as \codeid{PolyDet int size(PolyDet List<T> list)}. Invoking this method
with an argument of type \codeid{OrderNonDet List<Det String>} will result in \<PolyDet> getting resolved to \<OrderNonDet> which is invalid
on the return type \<int>. We circumvent this scenario by introducing two variants to the \<PolyDet> qualifier, \PolyDetUp
and \PolyDetDown.

@@

The semantics of these variants are as follows:
\begin{itemize}
    \item \PolyDetUp: Same as \<PolyDet> except when it resolves to \<OrderNonDet>, it is replaced with \<NonDet>.
    \item \PolyDetDown: Same as \<PolyDet> except when it resolves to \<OrderNonDet>, it is replaced with \<Det>.
\end{itemize} 
The modifiers $\uparrow$ and $\downarrow$ are indicative of the subtyping relationship of \<OrderNonDet> with \<NonDet> and \<Det> respectively
in the determinism type hierarchy.

@@

In order to provide soundness guarantees, we must ensure that a collection that is
declared as \<Det> is not side-effected in non-deterministic ways. Consider the following qualifier 
on the List add method: \codeid{public void add(PolyDet ArrayList<E> this, PolyDet int index, E element)}.
Suppose a client calls this method as follows:
\begin{verbatim}
@Det ArrayList<@Det String> list;
@NonDet int random;
@Det String str;
list.add(random, str);
\end{verbatim}
This will result in the instantiation \codeid{public void add(@NonDet ArrayList<E> this, @NonDet int index, E element)} and
the method invocation will type check. This is problematic because even though \<list> was declared to be \<Det>,
the type checker allows the addition of an element at a \<NonDet> location violating our determinism guarantees.
We prevent this behavior by introducing another variant of \<PolyDet>, \<PolyDet("use")> that does not affect
type instantiations. This is especially useful in preventing methods from non-deterministically modifying the state
of a deterministic receiver. To have this effect, the library method \<ArrayList.add> is annotated as:
\codeid{public void add(PolyDet ArrayList<E> this, PolyDet("use") int index, E element)}.
Since \<PolyDet("use")> has no effect on type instantiations, calling this method on a \<Det List>
will result in the method declaration being resolved as \codeid{public void add(@Det ArrayList<E> this, @Det int index, E element)}
irrespective of the type of \<index>, thereby preventing undesired side-effects.

@@

    \caption{Improving precision}
    \label{fig-precision-rules}
\end{figure}
According to this rule, if the \<Set> receiver has type \codeid{@OrderNonDet Set<@Det List} and the argument to the \codeid{equals()} method is
also of the same type, the return type gets the qualifier \<Det>. Without this precision rule, the return type would have resolved to \<NonDet> which
is sound but imprecise. This rule is careful enough to not annotate the return type of \<equals()> as \<Det> if a type argument to either the
receiver or the argument happens to be an \codeid{OrderNonDet List} (An order nondeterministic list does not provide any guarantee on the
equality as it is dependent on the iteration order).

>>>>

\begin{wrapfigure}{R}{0.3\textwidth}
    \begin{center}
        \includegraphics[scale=0.37]{detHierarchy}
    \end{center}
    \caption{Determinism type qualifier hierarchy}
    \label{fig:determinism-hierarchy}
%\todo{Drop the at-signs (@).}
%\todo{The text is weirdly small for the box.
%  Probably redraw this.}
\end{wrapfigure}

@@

\Cref{type-validity} shows the type well formedness
%\todo{The caption says
%  ``type well formedness''.  Choose one term and use it consistently
%  throughout.}
rules of \ourTypeSystem. 
As stated in the rule \rulename{noncollection}, the types \<Det> and \<NonDet>
can be written on any non-collection Java basetype.

@@

When a user writes a polymorphic qualifier on a method signature or a type parameter,
it indicates that it could be instantiated
%\todo{Should all occurrences of ``resolve'' be ``can be instantiated''} 
to any type in the type system depending on how it is used.
In \theDeterminismChecker, we define a polymorphic qualifier \<PolyDet>.
One of the most common locations for a polymorphic qualifier is at a method signature.
Consider the following declaration of addition:
\begin{verbatim}
@PolyDet int plus(@PolyDet int a, @PolyDet int b) { return a+b; }
\end{verbatim}
\todo{It has the standard meaning.  There are two ways to interpret this.}
There are two valid instantiations for this declaration:
\begin{Verbatim}
  @NonDet int plus(@NonDet int a, @NonDet int b) { return a+b; }
  @Det int plus(@Det int a, @Det int b) { return a+b; }
\end{Verbatim}
This indicates that method \<foo> can be called with arguments having any of the \codeid{Det or NonDet} type qualifiers.
%   (\<@OrderNonDet> is not allowed because it is invalid on primitive types).
 \<PolyDet> gets instantiated to the least upper bound of
the actual types on arguments. For instance, if this method is called as \codeid{foo(@Det String arg1, @NonDet boolean arg2)}, \theDeterminismChecker instantiates the method declaration as \codeid{@NonDet int foo(@NonDet String param1, @NonDet boolean param2)}
causing the return type to have the type qualifier \<NonDet>.

@@

Recall that \theDeterminismChecker does not permit \<OrderNonDet> on a type
that isn't a collection.
As a consequence, resolution of polymorphic type qualifiers at method signatures could have undesirable effects.
Suppose a method declaration is annotated as \codeid{PolyDet int size(PolyDet List<T> list)}. Invoking this method
with an argument of type \codeid{OrderNonDet List<Det String>} will result in \<PolyDet> getting instantiated to \<OrderNonDet> which is invalid
on the return type \<int>. We circumvent this scenario by introducing two variants to the \<PolyDet> qualifier, \PolyDetUp
and \PolyDetDown.

@@

The semantics of these variants are as follows:
\begin{itemize}
    \item \PolyDetUp: Same as \<PolyDet> except when it is instantiated as \<OrderNonDet>, it is replaced with \<NonDet>.
    \item \PolyDetDown: Same as \<PolyDet> except when it is instantiated as \<OrderNonDet>, it is replaced with \<Det>.
\end{itemize} 
The modifiers $\uparrow$ and $\downarrow$ are indicative of the subtyping relationship of \<OrderNonDet> with \<NonDet> and \<Det> respectively
in the determinism type hierarchy.

@@

In order to provide soundness guarantees, we must ensure that a collection that is
declared as \<Det> is not side-effected in non-deterministic ways. Consider the following qualifier 
on the List add method: \codeid{public void add(PolyDet ArrayList<E> this, PolyDet int index, E element)}.
Suppose a client calls this method as follows:
\begin{verbatim}
@Det ArrayList<@Det String> list;
@NonDet int random;
@Det String str;
list.add(random, str);
\end{verbatim}
This will result in the instantiation \codeid{public void add(@NonDet ArrayList<E> this, @NonDet int index, E element)} and
the method invocation will type check. This is problematic because even though \<list> was declared to be \<Det>,
the type checker allows the addition of an element at a \<NonDet> location violating our determinism guarantees.
We prevent this behavior by introducing another variant of \<PolyDet>, \<PolyDet("use")> that does not affect
type instantiations. This is especially useful in preventing methods from non-deterministically modifying the state
of a deterministic receiver. To have this effect, the library method \<ArrayList.add> is annotated as:
\codeid{public void add(PolyDet ArrayList<E> this, PolyDet("use") int index, E element)}.
Since \<PolyDet("use")> has no effect on type instantiations, calling this method on a \<Det List>
will result in the method declaration being instantiated as \codeid{public void add(@Det ArrayList<E> this, @Det int index, E element)}
irrespective of the type of \<index>, thereby preventing undesired side-effects.

@@

    \caption{Improving precision}
    \label{fig-precision-rules}
\end{figure}
According to this rule, if the \<Set> receiver has type \codeid{@OrderNonDet Set<@Det List} and the argument to the \codeid{equals()} method is
also of the same type, the return type gets the qualifier \<Det>. Without this precision rule, the return type would have been instantiated as \<NonDet> which
is sound but imprecise. This rule is careful enough to not annotate the return type of \<equals()> as \<Det> if a type argument to either the
receiver or the argument happens to be an \codeid{OrderNonDet List} (An order nondeterministic list does not provide any guarantee on the
equality as it is dependent on the iteration order).
